---
title: "HRRI: Holobiont Redox Resilience Index — End-to-End Workflow"
author:
  - name: Mitra Ghotbi
    email: mitra.ghotbi@gmail.com
date: "`r Sys.Date()`"
package: HRRI
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 4.2
    dev: png
header-includes: |
  <style>
  /* ==========================================================================
     HRRI vignette theme — clean scientific
     Muted slate/earth palette, serif headings, generous whitespace.
     ========================================================================== */
  
  :root {
    --hrri-ink:        #1c2321;   /* body text */
    --hrri-ink-soft:   #4a5451;   /* secondary text */
    --hrri-rule:       #dde3e1;   /* hairlines */
    --hrri-bg:         #ffffff;
    --hrri-bg-soft:    #f7f9f8;   /* code / table zebra */
  
    --hrri-fe:         #8c4a2f;   /* iron   — rust        */
    --hrri-mn:         #6b5b8a;   /* mangan — violet      */
    --hrri-redox:      #2f6b6b;   /* redox  — deep teal   */
    --hrri-accent:     #2f6b6b;
    --hrri-warn:       #9a6a24;   /* caution ochre        */
  
    --hrri-serif:  "Source Serif 4", "Iowan Old Style", Palatino, Georgia, serif;
    --hrri-sans:   "Source Sans 3", -apple-system, BlinkMacSystemFont, "Segoe UI",
                   Helvetica, Arial, sans-serif;
    --hrri-mono:   "JetBrains Mono", "SF Mono", Menlo, Consolas, monospace;
  }
  
  /* ---------- page frame ---------------------------------------------------- */
  
  body {
    font-family: var(--hrri-sans);
    font-size: 16px;
    line-height: 1.65;
    color: var(--hrri-ink);
    background: var(--hrri-bg);
    max-width: 46rem;
    margin: 0 auto;
    padding: 3.5rem 1.5rem 6rem;
    -webkit-font-smoothing: antialiased;
    text-rendering: optimizeLegibility;
  }
  
  /* ---------- title block --------------------------------------------------- */
  
  h1.title {
    font-family: var(--hrri-serif);
    font-size: 2.15rem;
    font-weight: 600;
    line-height: 1.2;
    letter-spacing: -0.015em;
    color: var(--hrri-ink);
    margin: 0 0 0.4rem;
    padding-bottom: 1.1rem;
    border-bottom: 2px solid var(--hrri-redox);
  }
  
  h4.author, h4.date {
    font-family: var(--hrri-sans);
    font-size: 0.9rem;
    font-weight: 400;
    color: var(--hrri-ink-soft);
    margin: 0.35rem 0 0;
  }
  h4.author { margin-top: 1rem; }
  
  /* ---------- headings ------------------------------------------------------ */
  
  h1, h2, h3, h4 {
    font-family: var(--hrri-serif);
    font-weight: 600;
    color: var(--hrri-ink);
    letter-spacing: -0.01em;
  }
  
  h1 { font-size: 1.75rem; margin: 3.2rem 0 1rem; }
  
  h2 {
    font-size: 1.34rem;
    margin: 2.8rem 0 0.9rem;
    padding-bottom: 0.4rem;
    border-bottom: 1px solid var(--hrri-rule);
  }
  
  h3 {
    font-size: 1.1rem;
    margin: 2rem 0 0.6rem;
    color: var(--hrri-ink-soft);
  }
  
  h4 {
    font-size: 0.97rem;
    font-family: var(--hrri-sans);
    text-transform: uppercase;
    letter-spacing: 0.07em;
    color: var(--hrri-ink-soft);
    margin: 1.6rem 0 0.5rem;
  }
  
  /* ---------- body copy ----------------------------------------------------- */
  
  p { margin: 0 0 1.15rem; }
  
  a {
    color: var(--hrri-accent);
    text-decoration: none;
    border-bottom: 1px solid rgba(47, 107, 107, 0.32);
    transition: border-color 0.15s ease;
  }
  a:hover { border-bottom-color: var(--hrri-accent); }
  
  strong { font-weight: 650; color: var(--hrri-ink); }
  
  /* ---------- table of contents --------------------------------------------- */
  
  #TOC {
    background: var(--hrri-bg-soft);
    border: 1px solid var(--hrri-rule);
    border-radius: 6px;
    padding: 1.2rem 1.5rem;
    margin: 2rem 0 3rem;
    font-size: 0.9rem;
  }
  #TOC ul { list-style: none; padding-left: 1rem; margin: 0.3rem 0; }
  #TOC > ul { padding-left: 0; }
  #TOC a {
    border-bottom: none;
    color: var(--hrri-ink-soft);
    display: block;
    padding: 0.16rem 0;
  }
  #TOC a:hover { color: var(--hrri-accent); }
  
  /* ---------- code ---------------------------------------------------------- */
  
  code {
    font-family: var(--hrri-mono);
    font-size: 0.855em;
    background: var(--hrri-bg-soft);
    color: var(--hrri-fe);
    padding: 0.12em 0.36em;
    border-radius: 3px;
  }
  
  pre {
    font-family: var(--hrri-mono);
    font-size: 0.83rem;
    line-height: 1.55;
    background: var(--hrri-bg-soft);
    border: 1px solid var(--hrri-rule);
    border-left: 3px solid var(--hrri-redox);
    border-radius: 4px;
    padding: 1rem 1.15rem;
    margin: 1.3rem 0;
    overflow-x: auto;
  }
  pre code {
    background: none;
    color: var(--hrri-ink);
    padding: 0;
    font-size: inherit;
  }
  
  /* knitr output blocks — visually subordinate to input */
  pre:not([class]), pre.r + pre {
    border-left-color: var(--hrri-rule);
    color: var(--hrri-ink-soft);
    background: #fbfcfc;
  }
  
  /* syntax highlighting */
  code span.fu { color: var(--hrri-redox); font-weight: 600; }  /* function   */
  code span.st { color: var(--hrri-fe); }                        /* string     */
  code span.co { color: #7d8886; font-style: italic; }           /* comment    */
  code span.dv, code span.fl { color: var(--hrri-mn); }          /* numbers    */
  code span.kw { color: var(--hrri-ink); font-weight: 600; }     /* keyword    */
  
  /* ---------- tables -------------------------------------------------------- */
  
  table {
    width: 100%;
    border-collapse: collapse;
    font-size: 0.88rem;
    font-variant-numeric: tabular-nums;
    margin: 1.6rem 0;
  }
  
  th {
    font-family: var(--hrri-sans);
    font-size: 0.78rem;
    font-weight: 600;
    text-transform: uppercase;
    letter-spacing: 0.05em;
    color: var(--hrri-ink-soft);
    text-align: left;
    padding: 0.55rem 0.7rem;
    border-bottom: 2px solid var(--hrri-ink);
  }
  
  td {
    padding: 0.5rem 0.7rem;
    border-bottom: 1px solid var(--hrri-rule);
  }
  
  tbody tr:nth-child(even) { background: var(--hrri-bg-soft); }
  tbody tr:last-child td   { border-bottom: 2px solid var(--hrri-ink); }
  
  /* numeric columns right-align */
  td:not(:first-child) { text-align: right; }
  th:not(:first-child) { text-align: right; }
  
  /* ---------- figures ------------------------------------------------------- */
  
  img, .figure {
    max-width: 100%;
    height: auto;
    display: block;
    margin: 1.9rem auto;
  }
  
  p.caption, .figure p {
    font-size: 0.84rem;
    color: var(--hrri-ink-soft);
    text-align: left;
    margin-top: 0.7rem;
    padding-left: 0.9rem;
    border-left: 2px solid var(--hrri-rule);
    line-height: 1.5;
  }
  
  /* ---------- math ---------------------------------------------------------- */
  
  .math.display {
    margin: 1.8rem 0;
    padding: 1.1rem 0;
    overflow-x: auto;
    border-top: 1px solid var(--hrri-rule);
    border-bottom: 1px solid var(--hrri-rule);
  }
  
  /* ---------- callout boxes -------------------------------------------------
     Use in the .Rmd as:
       <div class="note">   ... </div>
       <div class="caution">... </div>
       <div class="method"> ... </div>
     -------------------------------------------------------------------------- */
  
  .note, .caution, .method {
    border-radius: 5px;
    padding: 1rem 1.25rem;
    margin: 1.7rem 0;
    font-size: 0.92rem;
    line-height: 1.6;
  }
  .note p:last-child, .caution p:last-child, .method p:last-child { margin-bottom: 0; }
  
  .note {
    background: #f2f7f7;
    border-left: 3px solid var(--hrri-redox);
  }
  
  .caution {
    background: #fbf6ec;
    border-left: 3px solid var(--hrri-warn);
  }
  
  .method {
    background: var(--hrri-bg-soft);
    border: 1px solid var(--hrri-rule);
    border-left: 3px solid var(--hrri-mn);
  }
  
  .note::before, .caution::before, .method::before {
    display: block;
    font-family: var(--hrri-sans);
    font-size: 0.71rem;
    font-weight: 700;
    text-transform: uppercase;
    letter-spacing: 0.1em;
    margin-bottom: 0.45rem;
  }
  .note::before    { content: "Note";        color: var(--hrri-redox); }
  .caution::before { content: "Caution";     color: var(--hrri-warn);  }
  .method::before  { content: "Method note"; color: var(--hrri-mn);    }
  
  /* ---------- blockquote ---------------------------------------------------- */
  
  blockquote {
    margin: 1.6rem 0;
    padding: 0.2rem 0 0.2rem 1.3rem;
    border-left: 3px solid var(--hrri-rule);
    color: var(--hrri-ink-soft);
    font-style: italic;
  }
  
  /* ---------- lists --------------------------------------------------------- */
  
  ul, ol { margin: 0 0 1.15rem; padding-left: 1.4rem; }
  li { margin-bottom: 0.38rem; }
  li::marker { color: var(--hrri-ink-soft); }
  
  /* ---------- horizontal rule ----------------------------------------------- */
  
  hr {
    border: none;
    border-top: 1px solid var(--hrri-rule);
    margin: 3rem 0;
  }
  
  /* ---------- print --------------------------------------------------------- */
  
  @media print {
    body { max-width: none; padding: 0; font-size: 10.5pt; }
    #TOC { display: none; }
    pre, table, .figure, .note, .caution, .method { page-break-inside: avoid; }
    h1, h2, h3 { page-break-after: avoid; }
    a { border-bottom: none; color: var(--hrri-ink); }
  }
  
  /* ---------- small screens ------------------------------------------------- */
  
  @media (max-width: 640px) {
    body { padding: 2rem 1.1rem 4rem; font-size: 15px; }
    h1.title { font-size: 1.7rem; }
    table { font-size: 0.8rem; }
    pre { font-size: 0.76rem; padding: 0.8rem; }
  }
  </style>
vignette: >
  %\VignetteIndexEntry{HRRI: Holobiont Redox Resilience Index — End-to-End Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}

library(HRRI)

knitr::opts_chunk$set(dev = "png", 
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  fig.align  = "center",
  dpi        = 200,
  out.width  = "100%",
  message    = FALSE,
  warning    = FALSE
)

## Consistent figure styling across the vignette -------------------------
if (requireNamespace("ggplot2", quietly = TRUE)) {
  old_theme <- ggplot2::theme_set(
    ggplot2::theme_minimal(base_size = 11) +
      ggplot2::theme(
        panel.grid.minor = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_line(linewidth = 0.3,
                                                 colour = "#dde3e1"),
        axis.title       = ggplot2::element_text(colour = "#4a5451"),
        axis.text        = ggplot2::element_text(colour = "#4a5451"),
        strip.text       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        plot.title       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        legend.position  = "bottom"
      )
  )
}

## Vignette-local adapter: retain every required design identifier even when
## a pipeline release returns only scores or a subset of identifier columns.
attach_hrri_ids <- function(scores, id) {
  scores <- as.data.frame(scores)
  id <- as.data.frame(id)
  keys <- c("plot", "depth", "plant_id", "time")

  if (anyDuplicated(names(scores)) || anyDuplicated(names(id))) {
    stop("Score and identifier tables must have unique column names.")
  }
  missing_keys <- setdiff(keys, names(id))
  if (length(missing_keys)) {
    stop("sim$id is missing: ", paste(missing_keys, collapse = ", "))
  }
  if (!nrow(id) || nrow(scores) != nrow(id)) {
    stop("Expected one pipeline score row per sim$id row.")
  }
  if (!"RRI" %in% names(scores) || !is.numeric(scores[["RRI"]])) {
    stop("Pipeline scores must contain a numeric RRI column.")
  }

  make_key <- function(x) {
    if (anyNA(x[keys])) stop("Observation identifiers cannot be missing.")
    values <- lapply(x[keys], as.character)
    if (any(vapply(values, function(v) any(grepl("\034", v, fixed = TRUE)),
                   logical(1)))) {
      stop("Observation identifiers contain the reserved key separator.")
    }
    do.call(paste, c(values, list(sep = "\034")))
  }

  id_key <- make_key(id)
  if (anyDuplicated(id_key)) {
    stop("sim$id must uniquely identify plot x depth x plant x time rows.")
  }

  ## Put scores in the original simulator order whenever usable keys exist.
  if (all(keys %in% names(scores))) {
    score_key <- make_key(scores)
    if (anyDuplicated(score_key)) stop("Duplicate observation keys in scores.")
    idx <- match(id_key, score_key)
    alignment <- "observation keys"
  } else if ("row_id" %in% names(id) && "row_id" %in% names(scores)) {
    id_row <- as.character(id[["row_id"]])
    score_row <- as.character(scores[["row_id"]])
    if (anyNA(id_row) || anyNA(score_row) ||
        anyDuplicated(id_row) || anyDuplicated(score_row)) {
      stop("row_id must be unique and nonmissing for key-based alignment.")
    }
    idx <- match(id_row, score_row)
    alignment <- "row_id"
  } else {
    ## Scores-only releases must preserve their input row order. Row count
    ## equality alone cannot prove this; do not sort either table beforehand.
    idx <- seq_len(nrow(id))
    alignment <- "input row order (pipeline contract)"
  }
  if (anyNA(idx)) stop("Some simulator identifiers have no matching scores.")
  scores <- scores[idx, , drop = FALSE]

  shared <- intersect(c(keys, "row_id"), intersect(names(scores), names(id)))
  for (nm in shared) {
    if (!identical(as.character(scores[[nm]]), as.character(id[[nm]]))) {
      stop("Conflicting or misaligned identifier column: ", nm)
    }
  }

  ## Rebuild identifiers once; never create depth.x/depth.y pairs.
  out <- cbind(id[keys], scores[setdiff(names(scores), keys)])
  rownames(out) <- NULL
  attr(out, "id_alignment") <- alignment
  out
}

finite_mean <- function(x) {
  x <- x[is.finite(x)]
  if (length(x)) mean(x) else NA_real_
}
```

# Introduction

The **HRRI** package implements exploratory, multi-domain diagnostics for describing how soil–plant–microbiome
systems buffer and recover from hydroclimatic redox disturbances (Ghotbi *et al.*,
2026). The theoretical framing distinguishes four properties. They are not all identifiable from a single observation curve:

| Property | Symbol | Interpretation |
|---|---|---|
| Capacity | $Q$ | Electron-accepting and electron-donating inventory available within the system (mmol e⁻ kg⁻¹) |
| Connectivity | $\alpha$ | Fraction of $Q$ electrochemically accessible to porewater |
| Kinetics | $k$ | Characteristic rate of electron exchange under physicochemical and biological constraints (h⁻¹) |
| Memory | $M$ | Legacy of prior disturbances retained through persistent biogeochemical, microbial and physiological states that influence future system responses |

These combine through the accessible-capacity formula:

$$C_{\mathrm{acc}} = \sum_i Q_i \cdot \alpha_i \cdot \left(1 - e^{-k_i \tau}\right)$$

and aggregate into the Holobiont Redox Resilience Index:

$$\mathrm{RRI}_{it} = w_P \cdot P_{it} + w_S \cdot S_{it} + w_M \cdot M_{it}$$

where $P$ (Physiology), $S$ (Soil), and $M$ (Microbial) are domain scores.

This vignette walks through the full workflow from in-silico data generation
to RRI computation, accessible-capacity estimation, and recovery-signature
metrics — all driven by `simulate_redox_holobiont()`.

# In-silico Data Generation

`simulate_redox_holobiont()` is the package's master simulator. It generates
synthetic longitudinal observations across all three holobiont domains.
<div class="caution">
Only Fe and Mn inventories carry closed-balance checks. Carbon, nitrogen,
sulfur and oxygen budgets are **not** closed, and all rate parameters are
illustrative rather than field-calibrated. Simulated output supports software
demonstration and falsifiable model checks, not empirical ecological inference.
</div>

```{r simulate}
library(HRRI)
packageVersion("HRRI")

## Compatibility shim -----------------------------------------------------
## rri_pipeline() is the convenience wrapper around rri_pipeline_st().
## If the *installed* HRRI predates the wrapper, define an equivalent local
## version so this vignette knits against either release. Reinstall the
## package (see README) to use the exported function directly.
if (!exists("rri_pipeline", mode = "function")) {
  message("Installed HRRI has no rri_pipeline(); using a vignette-local wrapper.")
  rri_pipeline <- function(dat = NULL, soil = NULL, plant = NULL,
                           micro = NULL, id = NULL,
                           domain_weights = c(Physio = 0.4, Soil = 0.35,
                                              Micro = 0.25), ...) {
    stopifnot(setequal(names(domain_weights),
                       c("Physio", "Soil", "Micro")))
    w <- domain_weights[c("Physio", "Soil", "Micro")]
    w <- w / sum(w)
    res <- rri_pipeline_st(
      ROS_flux = plant, Eh_stability = soil, micro_data = micro, id = id,
      w1 = unname(w[1]), w2 = unname(w[2]), w3 = unname(w[3]), ...
    )
    res$scores <- res$row_scores
    res
  }
}

## Reproducible 1-cycle flood-drain experiment
## n_plot=2, n_depth=2, n_plant=3, n_time=30 -> 360 rows
sim <- simulate_redox_holobiont(
  n_plot               = 2,
  n_depth              = 2,
  n_plant              = 3,
  n_time               = 30,
  p_micro              = 20,
  seed                 = 42,
  scenario             = "flood_drain",
  n_cycles             = 1,
  disturbance_strength = 0.70,
  history_strength     = 0.55,
  decoupling           = 0.20
)

## Top-level structure
names(sim)
nrow(sim$id)                   # one row per plot × depth × plant × time
```

## Design identifiers

```{r design}
head(sim$id[, c("plot","depth","plant_id","time","cycle","phase","WFPS")])
```

## Soil geochemical outputs

```{r soil}
head(sim$soil_data[, c("EAC","EDC","Cacc_EAC","Cacc_total","Cacc_fraction",
                        "FeIII_poor_crystalline_mmol_kg",
                        "FeII_mmol_kg","Eh","pH")])
```

## Fe mass-balance verification

```{r conservation}
## Maximum absolute error should be < 0.01 mmol kg-1
sim$conservation_checks
```

## Plant physiology

```{r plant}
head(sim$plant_data[, c("SPAD","FvFm","ROL","ROS_load","aerenchyma")])
```

## Microbial functional genes

```{r genes}
## 18 genes spanning Fe-cycling, denitrification, nitrification,
## methanogenesis, and sulfur cycling
colnames(sim$micro_gene_abundance)
summary(sim$micro_gene_abundance[, "mcrA"])   # methanogenesis gene
```

# Accessible-Capacity Estimation

`rri_accessible_capacity()` computes $C_{\mathrm{acc}}$ for arbitrary mineralogical
reservoirs with explicitly supplied accessibility and exchange-rate parameters.
The values below are illustrative model inputs, not estimates or validated
literature defaults.

```{r cacc}
## Subset one plot-depth unit for illustration
idx <- sim$id$plot == "P1" & sim$id$depth == "D1" & sim$id$plant_id == "Plant1"
sdf <- sim$soil_data[idx, ]

## Define reservoir specifications
## (Q_col names must match columns in sdf)
res_spec <- list(
  reactive_FeIII = list(
    Q_col = "FeIII_poor_crystalline_mmol_kg",
    alpha = "alpha_accept",   # column name: per-row connectivity
    k     = "k_accept_h",     # column name: per-row kinetics
    type  = "EAC"
  ),
  crystalline_FeIII = list(
    Q_col = "FeIII_crystalline_mmol_kg",
    alpha = 0.20,             # attenuated connectivity for crystalline phases
    k     = 0.008,            # h-1: slow exchange (goethite/hematite)
    type  = "EAC"
  ),
  FeII_pool = list(
    Q_col = "FeII_mmol_kg",
    alpha = "alpha_donate",
    k     = "k_donate_h",
    type  = "EDC"
  )
)

## tau = 24 h (diurnal event timescale)
cap <- rri_accessible_capacity(sdf, res_spec, tau = 24,
                                normalise = FALSE, return_components = TRUE)

## Per-component summary (returned because return_components = TRUE)
cap$components

## Mean accessible vs. total inventory
cat("Mean Cacc_raw:", mean(cap$cacc_raw, na.rm=TRUE), "mmol e- kg-1\n")
cat("Mean fraction :", mean(cap$cacc_fraction, na.rm=TRUE), "\n")
if ("ck_limited" %in% names(cap) && length(cap$ck_limited)) {
  cat("CK-limited rows:", sum(cap$ck_limited, na.rm=TRUE),
      "/", sum(!is.na(cap$ck_limited)), "classified rows\n")
} else {
  cat("CK-limited classification is not returned by this HRRI version.\n")
}
```

## Effect of event timescale τ

```{r tau_sweep}
tau_vals <- c(1, 6, 24, 72, 168, 720)   # 1 h to 30 d
cacc_tau <- sapply(tau_vals, function(tt) {
  r <- rri_accessible_capacity(sdf, res_spec, tau = tt, normalise = FALSE)
  mean(r$cacc_raw, na.rm = TRUE)
})
data.frame(tau_h = tau_vals, Cacc_mean = round(cacc_tau, 2))
```

# RRI Pipeline

`rri_pipeline()` integrates available observed domains into an exploratory
composite. The simulator's hidden architecture columns are excluded. We use
explicit measured anchors to orient otherwise arbitrary PCA axes.

```{r rri_pipeline}
rri_out <- rri_pipeline(
  plant        = sim$ROS_flux,
  soil         = sim$Eh_stability,
  micro        = log1p(sim$micro_gene_abundance),
  id           = sim$id,
  mode         = "snapshot",
  scaling      = "pnorm",
  direction_anchor_phys = "FvFm",
  direction_anchor_soil = "Eh",
  direction_anchor_micro = "mtrA",
  domain_weights = c(Physio=0.35, Soil=0.40, Micro=0.25)
)

## Align once and reuse this identifier-complete table downstream.
rri_scored <- attach_hrri_ids(rri_out$row_scores, sim$id)
attr(rri_scored, "id_alignment")
summary(rri_scored$RRI)
head(rri_scored[, c("plot", "depth", "plant_id", "time",
                    "RRI", "Physio", "Soil", "Micro")])
```

<div class="method">
Alignment uses complete observation keys, or a shared unique `row_id` when
available. If neither is returned the pipeline must preserve input row order —
matching row counts alone do **not** establish alignment. Shared identifier
columns are checked for conflicts rather than silently overwritten.
</div>

## Agreement with the prescribed target

A single pooled correlation is the wrong summary here, for two reasons.

First, repeated rows and plant-depth trajectories within a plot are not
independent experimental units. The small workflow example has only two plots.
For the agreement illustration below we therefore generate a separate set of
24 plots, with 144 nested trajectories and 5760 rows, matching the paper design.
Intervals resample whole plots; treating rows as independent generally
understates uncertainty when within-plot dependence matters.

Second, Pearson's $r$ measures *association*, not *agreement*. A score equal to
twice the target plus a constant correlates with it perfectly while matching it
nowhere. Lin's concordance correlation coefficient penalises departure from the
1:1 line and is the quantity that belongs beside it.

`rri_accuracy()` reports both, with intervals obtained by resampling whole
independent plots rather than rows.

```{r validation}
cat("HRRI: starting the 24-plot agreement example (100 bootstrap resamples).\n",
    file = stderr())
acc_sim <- simulate_redox_holobiont(
  n_plot=24, n_depth=2, n_plant=3, n_time=40,
  seed=4096, scenario="flood_drain", disturbance_strength=.70,
  n_cycles=2L, disturbance_center=NULL, disturbance_width=.08)
acc_res <- rri_pipeline_st(
  ROS_flux=acc_sim$plant_data, Eh_stability=acc_sim$Eh_stability,
  micro_data=log1p(acc_sim$micro_gene_abundance), id=acc_sim$id,
  time_col="time", group_cols=c("plot","depth","plant_id"), mode="snapshot",
  direction_anchor_phys="FvFm", direction_anchor_soil="Eh", direction_anchor_micro="mtrA")
acc <- rri_accuracy(
  score=acc_res$row_scores$RRI, target=acc_sim$latent_truth,
  cluster=acc_sim$id$plot, n_boot=100, n_perm=0, seed=20260913)
cat("HRRI: agreement calculations finished; preparing figures.\n", file = stderr())
acc
```

This fast documentation example uses 100 bootstrap resamples. The complete
paper-export example uses 2000. The intervals condition on supplied fitted
scores and do not include pipeline refitting. No permutation P value is quoted:
exchangeability and time alignment require a separately justified design.


<div class="method">
`effective_n` is an ICC-based heuristic, not the effective sample size for
every statistic. Precision depends on the sampling design and estimand. Compare
the actual bootstrap intervals and retain the plot as the independent unit.
</div>

Splitting the error says which kind of disagreement is present, and the three
components sum to the mean squared error exactly.

```{r validation_decomp}
acc$decomposition[, c("component", "percent")]

## Exactness check: the residual is numerical noise, not a rounding allowance.
c(mse      = attr(acc$decomposition, "mse"),
  residual = attr(acc$decomposition, "residual"))
```

Large squared bias is a systematic offset, removable by recentring. Large
variance mismatch means the score is flatter or more volatile than the target.
Large lack of correlation means the score does not track the target's pattern,
and no rescaling will repair it.

`plot_rri_accuracy()` draws the same four questions as one figure.

```{r validation_figure, fig.width=7.4, fig.height=6.5, out.width="100%", fig.alt="Agreement with the prescribed target"}
# Use the installed namespace, avoiding a stale function in the workspace.
accuracy_plot <- HRRI::plot_rri_accuracy
plot_formals <- names(formals(accuracy_plot))
plot_args <- list(acc = acc, score_label = "RRI",
                  target_label = "Prescribed target",
                  base_size = 9, show_clusters = FALSE)
# Older HRRI builds do not accept these presentation arguments.
if ("cluster_label" %in% plot_formals) plot_args$cluster_label <- "Plots"
if ("style" %in% plot_formals) plot_args$style <- "paper"
accuracy_figure <- do.call(accuracy_plot, plot_args)
# Without optional patchwork, HRRI returns a named list of plots.
if (inherits(accuracy_figure, c("ggplot", "patchwork"))) {
  print(accuracy_figure)
} else {
  for (panel in accuracy_figure) print(panel)
}
```

<div class="note">
**Reading it.** The calibration panel compares the scores with the prescribed
target and the 1:1 line; correlation alone does not establish agreement. Open
points represent plot means. The difference panel shows score minus target
against their average, with descriptive limits for plot-mean differences, not
individual observations. The precision panel compares row resampling with
whole-plot resampling. Its intervals condition on supplied score-target pairs
and do not refit the scoring pipeline. The error panel partitions mean squared
error into squared bias, variance mismatch and lack of correlation.

The updated package uses the paper layout; older builds use their diagnostic
layout. The compatibility code changes presentation arguments only. Install the
updated package to reproduce the paper's figure styling.

**What it does not show.** None of the four panels speaks to out-of-sample
performance. The target is prescribed by the same simulator that produced the
inputs, so a tight panel A means the estimator is self-consistent, not that it
would recover an unobserved field quantity.
</div>

<div class="caution">
`latent_truth` and `RRI` are produced by the same generator. Everything above
therefore quantifies **internal consistency** — whether the estimator recovers
the target its own simulator prescribed. It is not independent predictive
validation, and it is not evidence that the individual latent parameters have
been identified. An empirical claim requires a target measured independently of
the score, replicated across independent experimental units.
</div>

# Recovery Signatures

`rri_recovery_metrics()` summarizes the RRI trajectory around a specified
disturbance window. The output fields depend on the installed package version;
the complete returned table is displayed below without assuming legacy names.
A score minimum or persistent departure alone does not establish biochemical
pathway truncation or alternative routing. Those interpretations require
independent process evidence.

The function requires **one row per time point per group**. Because the
pipeline returns one row per plant, we must first average RRI across plants
within each plot × depth × time cell before calling recovery metrics.

```{r recovery}
## Step 1 — use the aligned score table created in the pipeline chunk.
## Step 2 — aggregate to one row per plot × depth × time (mean over plants).
## The data-frame method avoids formula-level complete-case filtering.
## An all-missing group remains NA; it is not replaced by zero.
rri_agg <- stats::aggregate(
  x = rri_scored["RRI"],
  by = rri_scored[c("plot", "depth", "time")],
  FUN = finite_mean
)
rri_agg <- rri_agg[order(rri_agg$plot, rri_agg$depth, rri_agg$time), ,
                   drop = FALSE]
rownames(rri_agg) <- NULL
stopifnot(!anyDuplicated(rri_agg[c("plot", "depth", "time")]))

## Step 3 — extract recovery signatures.
## Pass rri_agg directly (group columns are already inside it; no id= needed).
metrics <- rri_recovery_metrics(
  res           = rri_agg,
  time_col      = "time",
  group_cols    = c("plot","depth"),
  perturb_start = 8,
  perturb_end   = 18,
  rri_col       = "RRI",
  forcing_col   = NULL   # no measured forcing supplied
)

if (!is.data.frame(metrics) || nrow(metrics) == 0L) {
  stop("rri_recovery_metrics() returned no nonempty recovery data frame.")
}
names(metrics)
metrics
```

With `forcing_col = NULL`, a returned hysteresis-related field must be
interpreted according to the installed function's documented definition; it
does not demonstrate a measured forcing–response loop. The limits 8 and 18
are example analysis boundaries, in the units of `time`; verify them against
the simulator's event schedule before interpreting recovery rates.

# Disturbance-History Sensitivity

Repeated-cycle simulations can test consequences of this simulator's stated
rules. They do not establish a universal mineralogical ratchet. In this version,
crystallisation is continuous rather than restricted to reoxidation events, and
end-state EAC is not constrained to decline monotonically with cycle count.

```{r history_sensitivity}
history <- do.call(rbind, lapply(1:4, function(nc) {
  z <- simulate_redox_holobiont(n_plot=1, n_depth=1, n_plant=2,
    n_time=30, p_micro=5, seed=99, n_cycles=nc,
    disturbance_strength=0.70)
  keep <- z$id$plant_id=="Plant1"
  data.frame(n_cycles=nc,
    EAC_end=tail(z$soil_data$EAC[keep],1),
    memory_end=tail(z$latent_state$memory[keep],1))
}))
history
```

Interpret the direction and magnitude as a model sensitivity result. An
empirical claim about hydrological memory requires independent observations.

```{r restore-theme, include=FALSE}
## theme_set() changes state that persists for the rest of the session.
## Vignettes build in their own process so nothing outside is affected, but
## restoring is the same courtesy CRAN asks for with par() and options().
if (exists("old_theme")) ggplot2::theme_set(old_theme)
```

# Publication figures

`vignette("HRRI_paper_figures")` contains all six manuscript figures, their
function mapping and interpretation. The original gallery and this workflow
remain available. All figures can be generated entirely in R; export vector PDF
or editable-text SVG at the supplied physical dimensions. The framework is a
schematic, the identifiability illustration is analytical, and the gallery and
agreement panels use different explicitly declared simulations.

# Session Information

```{r sessionInfo}
sessionInfo()
```

# References

## Published methods

Keiluweit, M., Wanzek, T., Kleber, M., Nico, P., & Fendorf, S. (2017).
Anaerobic microsites have an unaccounted role in soil carbon stabilization.
*Nature Communications*, **8**, 1771. <https://doi.org/10.1038/s41467-017-01406-6>

Klüpfel, L., Piepenbrock, A., Kappler, A., & Sander, M. (2014). Humic
substances as fully regenerable electron acceptors in recurrently anoxic
environments. *Nature Geoscience*, **7**, 195–200.
<https://doi.org/10.1038/ngeo2084>

Kobayashi, K., & Salam, M. U. (2000). Comparing simulated and measured values
using mean squared deviation and its components. *Agronomy Journal*, **92**,
345–352. <https://doi.org/10.2134/agronj2000.922345x>

Lin, L. I. (1989). A concordance correlation coefficient to evaluate
reproducibility. *Biometrics*, **45**, 255–268.
<https://doi.org/10.2307/2532051>

Sander, M., Hofstetter, T. B., & Gorski, C. A. (2015). Electrochemical analyses
of redox-active iron minerals: a review of nonmediated and mediated approaches.
*Environmental Science & Technology*, **49**, 5862–5878.
<https://doi.org/10.1021/acs.est.5b00006>

Thompson, A., Chadwick, O. A., Rancourt, D. G., & Chorover, J. (2006).
Iron-oxide crystallinity increases during soil redox oscillations.
*Geochimica et Cosmochimica Acta*, **70**, 1710–1727.
<https://doi.org/10.1016/j.gca.2005.12.005>

## Companion manuscripts

These describe the framework this package implements. None is published and two
are under review; the entries are provisional and should be replaced with the
published versions.

Ghotbi, M., Ghotbi, M., Komluski, J., & Holtgrewe-Stukenbrock, E. H. HRRI:
direction-aware diagnostics for soil–plant–microbiome redox recovery across
hydroclimatic disturbances. *In preparation.*

Ghotbi, M., Kolody, B. C., Ghotbi, M., & Holtgrewe-Stukenbrock, E. A Theory of
Hydroclimatic Redox Resilience. *Submitted to Communications Earth &
Environment.* — the source of the capacity, connectivity, kinetics and memory
decomposition used throughout this vignette.

Ghotbi, M., Ghotbi, M., Mühling, K. H., & Stukenbrock, E. H. Rhizosphere redox
recovery after hydrological disturbances: mechanisms across the
soil–plant–microbiome continuum. *Submitted to Soil Biology & Biochemistry.*
