---
title: "HRRI: Holobiont Redox Resilience Index — End-to-End Workflow"
author:
  - name: Mitra Ghotbi
    email: mitra.ghotbi@gmail.com
date: "`r Sys.Date()`"
package: HRRI
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    css: css/hrri.css
    fig_width: 7
    fig_height: 4.2
    dev: png
vignette: >
  %\VignetteIndexEntry{HRRI: Holobiont Redox Resilience Index — End-to-End Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  fig.align  = "center",
  dpi        = 150,
  out.width  = "100%",
  message    = FALSE,
  warning    = FALSE
)

## Consistent figure styling across the vignette -------------------------
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(
    ggplot2::theme_minimal(base_size = 11) +
      ggplot2::theme(
        panel.grid.minor = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_line(linewidth = 0.3,
                                                 colour = "#dde3e1"),
        axis.title       = ggplot2::element_text(colour = "#4a5451"),
        axis.text        = ggplot2::element_text(colour = "#4a5451"),
        strip.text       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        plot.title       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        legend.position  = "bottom"
      )
  )
}

## Vignette-local adapter: retain every required design identifier even when
## a pipeline release returns only scores or a subset of identifier columns.
attach_hrri_ids <- function(scores, id) {
  scores <- as.data.frame(scores)
  id <- as.data.frame(id)
  keys <- c("plot", "depth", "plant_id", "time")

  if (anyDuplicated(names(scores)) || anyDuplicated(names(id))) {
    stop("Score and identifier tables must have unique column names.")
  }
  missing_keys <- setdiff(keys, names(id))
  if (length(missing_keys)) {
    stop("sim$id is missing: ", paste(missing_keys, collapse = ", "))
  }
  if (!nrow(id) || nrow(scores) != nrow(id)) {
    stop("Expected one pipeline score row per sim$id row.")
  }
  if (!"RRI" %in% names(scores) || !is.numeric(scores[["RRI"]])) {
    stop("Pipeline scores must contain a numeric RRI column.")
  }

  make_key <- function(x) {
    if (anyNA(x[keys])) stop("Observation identifiers cannot be missing.")
    values <- lapply(x[keys], as.character)
    if (any(vapply(values, function(v) any(grepl("\034", v, fixed = TRUE)),
                   logical(1)))) {
      stop("Observation identifiers contain the reserved key separator.")
    }
    do.call(paste, c(values, list(sep = "\034")))
  }

  id_key <- make_key(id)
  if (anyDuplicated(id_key)) {
    stop("sim$id must uniquely identify plot x depth x plant x time rows.")
  }

  ## Put scores in the original simulator order whenever usable keys exist.
  if (all(keys %in% names(scores))) {
    score_key <- make_key(scores)
    if (anyDuplicated(score_key)) stop("Duplicate observation keys in scores.")
    idx <- match(id_key, score_key)
    alignment <- "observation keys"
  } else if ("row_id" %in% names(id) && "row_id" %in% names(scores)) {
    id_row <- as.character(id[["row_id"]])
    score_row <- as.character(scores[["row_id"]])
    if (anyNA(id_row) || anyNA(score_row) ||
        anyDuplicated(id_row) || anyDuplicated(score_row)) {
      stop("row_id must be unique and nonmissing for key-based alignment.")
    }
    idx <- match(id_row, score_row)
    alignment <- "row_id"
  } else {
    ## Scores-only releases must preserve their input row order. Row count
    ## equality alone cannot prove this; do not sort either table beforehand.
    idx <- seq_len(nrow(id))
    alignment <- "input row order (pipeline contract)"
  }
  if (anyNA(idx)) stop("Some simulator identifiers have no matching scores.")
  scores <- scores[idx, , drop = FALSE]

  shared <- intersect(c(keys, "row_id"), intersect(names(scores), names(id)))
  for (nm in shared) {
    if (!identical(as.character(scores[[nm]]), as.character(id[[nm]]))) {
      stop("Conflicting or misaligned identifier column: ", nm)
    }
  }

  ## Rebuild identifiers once; never create depth.x/depth.y pairs.
  out <- cbind(id[keys], scores[setdiff(names(scores), keys)])
  rownames(out) <- NULL
  attr(out, "id_alignment") <- alignment
  out
}

finite_mean <- function(x) {
  x <- x[is.finite(x)]
  if (length(x)) mean(x) else NA_real_
}
```

# Introduction

The **HRRI** package implements exploratory, multi-domain diagnostics for describing how soil–plant–microbiome
systems buffer and recover from hydroclimatic redox disturbances (Ghotbi *et al.*,
2026). The theoretical framing distinguishes four properties. They are not all identifiable from a single observation curve:

| Property | Symbol | Interpretation |
|---|---|---|
| Capacity | $Q$ | Total electron-accepting/donating inventory (mmol e⁻ kg⁻¹) |
| Connectivity | $\alpha$ | Fraction of $Q$ electrochemically accessible to porewater |
| Kinetics | $k$ | Exchange rate (h⁻¹); encodes mineralogical crystallinity |
| Memory | $M$ | Legacy of prior disturbances; encoded in Fe-phase composition |

These combine through the accessible-capacity formula:

$$C_{\rm acc} = \sum_i Q_i \cdot \alpha_i \cdot \left(1 - e^{-k_i \tau}\right)$$

and aggregate into the Holobiont Redox Resilience Index:

$$\mathrm{RRI}_{it} = w_P \cdot P_{it} + w_S \cdot S_{it} + w_M \cdot M_{it}$$

where $P$ (Physiology), $S$ (Soil), and $M$ (Microbial) are domain scores.

This vignette walks through the full workflow from in-silico data generation
to RRI computation, accessible-capacity estimation, and recovery-signature
metrics — all driven by `simulate_redox_holobiont()`.

# In-silico Data Generation

`simulate_redox_holobiont()` is the package's master simulator. It generates
synthetic longitudinal observations across all three holobiont domains.
<div class="caution">
Only Fe and Mn inventories carry closed-balance checks. Carbon, nitrogen,
sulfur and oxygen budgets are **not** closed, and all rate parameters are
illustrative rather than field-calibrated. Simulated output supports software
demonstration and falsifiable model checks, not empirical ecological inference.
</div>

```{r simulate}
library(HRRI)
packageVersion("HRRI")

## Compatibility shim -----------------------------------------------------
## rri_pipeline() is the convenience wrapper around rri_pipeline_st().
## If the *installed* HRRI predates the wrapper, define an equivalent local
## version so this vignette knits against either release. Reinstall the
## package (see README) to use the exported function directly.
if (!exists("rri_pipeline", mode = "function")) {
  message("Installed HRRI has no rri_pipeline(); using a vignette-local wrapper.")
  rri_pipeline <- function(dat = NULL, soil = NULL, plant = NULL,
                           micro = NULL, id = NULL,
                           domain_weights = c(Physio = 0.4, Soil = 0.35,
                                              Micro = 0.25), ...) {
    stopifnot(setequal(names(domain_weights),
                       c("Physio", "Soil", "Micro")))
    w <- domain_weights[c("Physio", "Soil", "Micro")]
    w <- w / sum(w)
    res <- rri_pipeline_st(
      ROS_flux = plant, Eh_stability = soil, micro_data = micro, id = id,
      w1 = unname(w[1]), w2 = unname(w[2]), w3 = unname(w[3]), ...
    )
    res$scores <- res$row_scores
    res
  }
}

## Reproducible 1-cycle flood-drain experiment
## n_plot=2, n_depth=2, n_plant=3, n_time=30 -> 360 rows
sim <- simulate_redox_holobiont(
  n_plot               = 2,
  n_depth              = 2,
  n_plant              = 3,
  n_time               = 30,
  p_micro              = 20,
  seed                 = 42,
  scenario             = "flood_drain",
  n_cycles             = 1,
  disturbance_strength = 0.70,
  history_strength     = 0.55,
  decoupling           = 0.20
)

## Top-level structure
names(sim)
nrow(sim$id)                   # one row per plot × depth × plant × time
```

## Design identifiers

```{r design}
head(sim$id[, c("plot","depth","plant_id","time","cycle","phase","WFPS")])
```

## Soil geochemical outputs

```{r soil}
head(sim$soil_data[, c("EAC","EDC","Cacc_EAC","Cacc_total","Cacc_fraction",
                        "FeIII_poor_crystalline_mmol_kg",
                        "FeII_mmol_kg","Eh","pH")])
```

## Fe mass-balance verification

```{r conservation}
## Maximum absolute error should be < 0.01 mmol kg-1
sim$conservation_checks
```

## Plant physiology

```{r plant}
head(sim$plant_data[, c("SPAD","FvFm","ROL","ROS_load","aerenchyma")])
```

## Microbial functional genes

```{r genes}
## 18 genes spanning Fe-cycling, denitrification, nitrification,
## methanogenesis, and sulfur cycling
colnames(sim$micro_gene_abundance)
summary(sim$micro_gene_abundance[, "mcrA"])   # methanogenesis gene
```

# Accessible-Capacity Estimation

`rri_accessible_capacity()` computes $C_{\rm acc}$ for arbitrary mineralogical
reservoirs with explicitly supplied accessibility and exchange-rate parameters.
The values below are illustrative model inputs, not estimates or validated
literature defaults.

```{r cacc}
## Subset one plot-depth unit for illustration
idx <- sim$id$plot == "P1" & sim$id$depth == "D1" & sim$id$plant_id == "Plant1"
sdf <- sim$soil_data[idx, ]

## Define reservoir specifications
## (Q_col names must match columns in sdf)
res_spec <- list(
  reactive_FeIII = list(
    Q_col = "FeIII_poor_crystalline_mmol_kg",
    alpha = "alpha_accept",   # column name: per-row connectivity
    k     = "k_accept_h",     # column name: per-row kinetics
    type  = "EAC"
  ),
  crystalline_FeIII = list(
    Q_col = "FeIII_crystalline_mmol_kg",
    alpha = 0.20,             # attenuated connectivity for crystalline phases
    k     = 0.008,            # h-1: slow exchange (goethite/hematite)
    type  = "EAC"
  ),
  FeII_pool = list(
    Q_col = "FeII_mmol_kg",
    alpha = "alpha_donate",
    k     = "k_donate_h",
    type  = "EDC"
  )
)

## tau = 24 h (diurnal event timescale)
cap <- rri_accessible_capacity(sdf, res_spec, tau = 24,
                                normalise = FALSE, return_components = TRUE)

## Per-component summary (returned because return_components = TRUE)
cap$components

## Mean accessible vs. total inventory
cat("Mean Cacc_raw:", mean(cap$cacc_raw, na.rm=TRUE), "mmol e- kg-1\n")
cat("Mean fraction :", mean(cap$cacc_fraction, na.rm=TRUE), "\n")
if ("ck_limited" %in% names(cap) && length(cap$ck_limited)) {
  cat("CK-limited rows:", sum(cap$ck_limited, na.rm=TRUE),
      "/", sum(!is.na(cap$ck_limited)), "classified rows\n")
} else {
  cat("CK-limited classification is not returned by this HRRI version.\n")
}
```

## Effect of event timescale τ

```{r tau_sweep}
tau_vals <- c(1, 6, 24, 72, 168, 720)   # 1 h to 30 d
cacc_tau <- sapply(tau_vals, function(tt) {
  r <- rri_accessible_capacity(sdf, res_spec, tau = tt, normalise = FALSE)
  mean(r$cacc_raw, na.rm = TRUE)
})
data.frame(tau_h = tau_vals, Cacc_mean = round(cacc_tau, 2))
```

# RRI Pipeline

`rri_pipeline()` integrates available observed domains into an exploratory
composite. The simulator's hidden architecture columns are excluded. We use
explicit measured anchors to orient otherwise arbitrary PCA axes.

```{r rri_pipeline}
rri_out <- rri_pipeline(
  plant        = sim$ROS_flux,
  soil         = sim$Eh_stability,
  micro        = log1p(sim$micro_gene_abundance),
  id           = sim$id,
  mode         = "snapshot",
  scaling      = "pnorm",
  direction_anchor_phys = "FvFm",
  direction_anchor_soil = "Eh",
  direction_anchor_micro = "mtrA",
  domain_weights = c(Physio=0.35, Soil=0.40, Micro=0.25)
)

## Align once and reuse this identifier-complete table downstream.
rri_scored <- attach_hrri_ids(rri_out$row_scores, sim$id)
attr(rri_scored, "id_alignment")
summary(rri_scored$RRI)
head(rri_scored[, c("plot", "depth", "plant_id", "time",
                    "RRI", "Physio", "Soil", "Micro")])
```

<div class="method">
Alignment uses complete observation keys, or a shared unique `row_id` when
available. If neither is returned the pipeline must preserve input row order —
matching row counts alone do **not** establish alignment. Shared identifier
columns are checked for conflicts rather than silently overwritten.
</div>

## Correlation with latent truth

```{r validation}
## rri_scored is aligned to sim$id, and hence to its latent_truth vector.
truth <- sim$latent_truth
if (!is.numeric(truth) || length(truth) != nrow(rri_scored)) {
  stop("latent_truth must be a numeric vector with one value per sim$id row.")
}
ok <- is.finite(rri_scored$RRI) & is.finite(truth)
if (sum(ok) >= 3L && stats::sd(rri_scored$RRI[ok]) > 0 &&
    stats::sd(truth[ok]) > 0) {
  r_val <- stats::cor(rri_scored$RRI[ok], truth[ok])
} else {
  r_val <- NA_real_
  cat("Correlation unavailable: too few finite pairs or a constant vector.\n")
}
cat("r(RRI, latent truth):", round(r_val, 3), "\n")
```

This is a within-simulation association, not independent predictive validation
or evidence that the individual latent parameters have been identified.

# Recovery Signatures

`rri_recovery_metrics()` summarizes the RRI trajectory around a specified
disturbance window. The output fields depend on the installed package version;
the complete returned table is displayed below without assuming legacy names.
A score minimum or persistent departure alone does not establish biochemical
pathway truncation or alternative routing. Those interpretations require
independent process evidence.

The function requires **one row per time point per group**. Because the
pipeline returns one row per plant, we must first average RRI across plants
within each plot × depth × time cell before calling recovery metrics.

```{r recovery}
## Step 1 — use the aligned score table created in the pipeline chunk.
## Step 2 — aggregate to one row per plot × depth × time (mean over plants).
## The data-frame method avoids formula-level complete-case filtering.
## An all-missing group remains NA; it is not replaced by zero.
rri_agg <- stats::aggregate(
  x = rri_scored["RRI"],
  by = rri_scored[c("plot", "depth", "time")],
  FUN = finite_mean
)
rri_agg <- rri_agg[order(rri_agg$plot, rri_agg$depth, rri_agg$time), ,
                   drop = FALSE]
rownames(rri_agg) <- NULL
stopifnot(!anyDuplicated(rri_agg[c("plot", "depth", "time")]))

## Step 3 — extract recovery signatures.
## Pass rri_agg directly (group columns are already inside it; no id= needed).
metrics <- rri_recovery_metrics(
  res           = rri_agg,
  time_col      = "time",
  group_cols    = c("plot","depth"),
  perturb_start = 8,
  perturb_end   = 18,
  rri_col       = "RRI",
  forcing_col   = NULL   # no measured forcing supplied
)

if (!is.data.frame(metrics) || nrow(metrics) == 0L) {
  stop("rri_recovery_metrics() returned no nonempty recovery data frame.")
}
names(metrics)
metrics
```

With `forcing_col = NULL`, a returned hysteresis-related field must be
interpreted according to the installed function's documented definition; it
does not demonstrate a measured forcing–response loop. The limits 8 and 18
are example analysis boundaries, in the units of `time`; verify them against
the simulator's event schedule before interpreting recovery rates.

# Disturbance-History Sensitivity

Repeated-cycle simulations can test consequences of this simulator's stated
rules. They do not establish a universal mineralogical ratchet. In this version,
crystallisation is continuous rather than restricted to reoxidation events, and
end-state EAC is not constrained to decline monotonically with cycle count.

```{r history_sensitivity}
history <- do.call(rbind, lapply(1:4, function(nc) {
  z <- simulate_redox_holobiont(n_plot=1, n_depth=1, n_plant=2,
    n_time=30, p_micro=5, seed=99, n_cycles=nc,
    disturbance_strength=0.70)
  keep <- z$id$plant_id=="Plant1"
  data.frame(n_cycles=nc,
    EAC_end=tail(z$soil_data$EAC[keep],1),
    memory_end=tail(z$latent_state$memory[keep],1))
}))
history
```

Interpret the direction and magnitude as a model sensitivity result. An
empirical claim about hydrological memory requires independent observations.

# Session Information

```{r sessionInfo}
sessionInfo()
```

# References

Mitra Ghotbi¹˒², Marjan Ghotbi³, Marco Guerreiro1, Jovan komluski1, Eva Holtgrewe-Stukenbrock¹˒4 (2026).
HRRI: A direction-aware R framework for quantifying soil–plant–microbiome
redox resilience across hydroclimatic disturbance events.
Manuscript submitted.
