---
title: "HRRI: Reading the Figures — A Worked Example"
author:
  - name: Mitra Ghotbi
    email: mitra.ghotbi@gmail.com
date: "`r Sys.Date()`"
package: HRRI
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    css: css/hrri.css
    fig_width: 7
    fig_height: 4.4
    dev: png
vignette: >
  %\VignetteIndexEntry{HRRI: Reading the Figures — A Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.4,
  fig.align  = "center",
  dpi        = 150,
  out.width  = "100%",
  message    = FALSE,
  warning    = FALSE
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::theme_set(
    ggplot2::theme_minimal(base_size = 11) +
      ggplot2::theme(
        panel.grid.minor = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_line(linewidth = 0.3,
                                                 colour = "#dde3e1"),
        axis.title       = ggplot2::element_text(colour = "#4a5451"),
        axis.text        = ggplot2::element_text(colour = "#4a5451"),
        strip.text       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        plot.title       = ggplot2::element_text(face = "bold",
                                                 colour = "#1c2321"),
        legend.position  = "bottom"
      )
  )
}
```

# How to use this vignette

The companion vignette, `vignette("HRRI_workflow")`, shows *how to run* HRRI.
This one shows *how to read what comes out*.

We follow a single simulated experiment from beginning to end. Every figure is
produced from that one experiment, so the panels connect: the trajectory you
see in Section 4 is the same trajectory summarised in Section 6 and ranked in
Section 7.

Each figure is followed by two short blocks:

<div class="note">
**Reading it** — what the axes mean and what pattern to look for.
</div>

<div class="caution">
**What it does not show** — the inference the figure cannot support. HRRI is
deliberately conservative, and these limits are part of the method, not
disclaimers bolted on afterwards.
</div>

```{r libraries}
library(HRRI)
library(ggplot2)
packageVersion("HRRI")
```

# The experiment

One flood–drain cycle across two plots, two depths and three plants, observed
daily for 40 steps. The disturbance runs from day 12 to day 22.

```{r simulate}
PERTURB_START <- 12
PERTURB_END   <- 22

sim <- simulate_redox_holobiont(
  n_plot               = 2,
  n_depth              = 2,
  n_plant              = 3,
  n_time               = 40,
  p_micro              = 25,
  seed                 = 2026,
  scenario             = "flood_drain",
  n_cycles             = 1,
  disturbance_strength = 0.72,
  history_strength     = 0.55
)

nrow(sim$id)          # 2 x 2 x 3 x 40 = 480 observations
names(sim$latent_state)
```

<div class="caution">
`latent_state` is the simulator's ground truth. It exists so we can check
whether HRRI recovers what it is meant to recover. With real data there is no
such column, and nothing in the scoring path is allowed to read it.
</div>

# The four hidden states, made visible

HRRI rests on four controls that are never measured directly. In simulation we
can plot them, which is the clearest way to build intuition for what the index
is chasing.

```{r hidden-states, fig.height=5.2}
ls_df <- as.data.frame(sim$latent_state)
one   <- with(sim$id, plot == "P1" & depth == "D1" & plant_id == "Plant1")

hidden <- data.frame(
  time  = sim$id$time[one],
  value = c(
    ls_df$Q_accept[one],
    ls_df$alpha_accept[one],
    ls_df$k_accept_h[one],
    ls_df$memory[one]
  ),
  state = factor(
    rep(c("Capacity  (Q)", "Connectivity  (α)",
          "Kinetics  (k)", "Memory  (M)"), each = sum(one)),
    levels = c("Capacity  (Q)", "Connectivity  (α)",
               "Kinetics  (k)", "Memory  (M)")
  )
)

ggplot(hidden, aes(time, value)) +
  annotate("rect", xmin = PERTURB_START, xmax = PERTURB_END,
           ymin = -Inf, ymax = Inf, fill = "#2f6b6b", alpha = 0.10) +
  geom_line(colour = "#2f6b6b", linewidth = 0.7) +
  facet_wrap(~ state, scales = "free_y", ncol = 2) +
  labs(x = "Time (days)", y = NULL,
       title = "The four hidden states during one flood-drain cycle",
       subtitle = "Shaded band = disturbance window") +
  theme(legend.position = "none")
```

<div class="note">
**Reading it** — Capacity is the electron inventory and moves slowly.
Connectivity collapses during flooding as pore pathways close, then partially
reopens. Kinetics tracks the rate of exchange. Memory is the one that does not
come back: it ratchets upward and stays there. That asymmetry is the whole
point of the framework.
</div>

## Memory is a holobiont state

Since version 0.99.2, memory accumulates from four sources spanning all three
domains, not from iron chemistry alone. The components are exported so the
decomposition can be checked rather than assumed.

```{r memory-components, fig.height=3.6}
mem <- data.frame(
  time  = rep(sim$id$time[one], 3),
  value = c(ls_df$memory[one],
            ls_df$plant_legacy[one],
            ls_df$micro_legacy[one]),
  component = factor(
    rep(c("Memory (total)", "Plant legacy (aerenchyma)",
          "Microbial legacy (community)"), each = sum(one)),
    levels = c("Memory (total)", "Plant legacy (aerenchyma)",
               "Microbial legacy (community)")
  )
)

ggplot(mem, aes(time, value, colour = component, linetype = component)) +
  annotate("rect", xmin = PERTURB_START, xmax = PERTURB_END,
           ymin = -Inf, ymax = Inf, fill = "#9a6a24", alpha = 0.10) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c("#1c2321", "#8c4a2f", "#6b5b8a")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotdash")) +
  labs(x = "Time (days)", y = "Legacy (0-1)", colour = NULL, linetype = NULL,
       title = "Memory decomposed across the holobiont")
```

<div class="note">
**Reading it** — the microbial legacy rises quickly under sustained reduction
and relaxes slowly afterwards. That asymmetry is deliberate: a component that
tracked current conditions symmetrically would be a readout, not a memory.
</div>

<div class="caution">
**What it does not show** — that these are the *correct* weights. They are
illustrative (0.020 event load, 0.016 Fe ratchet, 0.012 plant, 0.012 microbial)
and are not calibrated to measured legacy effects in any field system.
</div>

# Scoring the three domains

```{r pipeline}
res <- rri_pipeline_st(
  ROS_flux     = sim$ROS_flux,
  Eh_stability = sim$Eh_stability,
  micro_data   = sim$micro_data,
  id           = sim$id,
  time_col     = "time",
  group_cols   = c("plot", "depth", "plant_id"),
  mode         = "snapshot",
  reducer      = "per_domain",
  scaling      = "pnorm",
  direction_anchor_phys  = "FvFm",
  direction_anchor_soil  = "Eh",
  direction_anchor_micro = "ASV1"
)

scored <- attach_hrri_ids(res$row_scores, sim$id)
attr(scored, "id_alignment")
summary(scored$RRI)
```

<div class="caution">
The `direction_anchor_*` arguments are not optional in practice. Latent axes
from PCA have arbitrary sign; without an anchor variable whose direction you
can justify, a high RRI could mean the opposite of what you intend. The
function warns when anchors are missing.
</div>

# One trajectory in context

```{r timeseries}
plot_rri_timeseries(
  sim, res,
  plot_id       = "P1",
  depth_id      = "D1",
  plant_id      = "Plant1",
  perturb_start = PERTURB_START,
  perturb_end   = PERTURB_END
)
```

<div class="note">
**Reading it** — Eh, accessible capacity and the composite RRI on a shared time
axis in separate panels, each in its own units. Look for whether RRI returns to
its pre-event level, and whether it returns at the same time as Eh. A gap
between the two is the interesting case.
</div>

<div class="caution">
**What it does not show** — panels are not placed on a common axis, because Eh
(mV) and RRI (dimensionless) are not commensurable. Visual co-movement is not
evidence of a mechanistic link.
</div>

# Where the domains sit relative to each other

```{r ternary, echo=TRUE, results="asis"}
## ggtern is a Suggests dependency and, at the time of writing, is not
## compatible with ggplot2 >= 3.5 (it errors on tern.axis.ticks.length.major).
## Render defensively so the vignette builds either way.
tern_ok <- requireNamespace("ggtern", quietly = TRUE) &&
           requireNamespace("viridis", quietly = TRUE)

if (tern_ok) {
  p_tern <- try(
    plot_RRI_ternary(res$row_scores_comp, point_size = 2.4,
                     show_centroid = TRUE),
    silent = TRUE
  )
  drawn <- !inherits(p_tern, "try-error") &&
           !inherits(try(print(p_tern), silent = TRUE), "try-error")
  if (!drawn) {
    cat("*The ternary plot could not be rendered: the installed **ggtern** is",
        "incompatible with this **ggplot2** version. The composition it would",
        "show is summarised numerically below.*\n\n")
  }
} else {
  cat("*Install **ggtern** and **viridis** to render this figure.*\n\n")
}
```

Whether or not the ternary renders, the same information is available directly
from the compositional table — each row sums to one across the three domains:

```{r ternary-numeric}
comp <- res$row_scores_comp[, c("Physio", "Soil", "Micro")]
round(colMeans(comp, na.rm = TRUE), 3)          # centroid
round(range(rowSums(comp, na.rm = TRUE)), 6)    # closure check: both 1
```

<div class="note">
**Reading it** — each point is one observation placed by the *relative* weight
of its Physiology, Soil and Microbial scores. Points near a vertex are
dominated by that domain. The white diamond is the centroid. Movement of the
cloud toward a vertex over an event means the domains are responding
unequally.
</div>

<div class="caution">
**What it does not show** — position is compositional, so it discards
magnitude. Two samples with very different absolute RRI sit at the same point
if their domain *ratios* match. Always read the ternary alongside Section 4.
</div>

# Domain-score state space

```{r state-space}
plot_rri_state_space(
  res,
  x_property = "Physio",
  y_property = "Soil",
  colour_by  = "RRI",
  group_cols = c("plot", "depth", "plant_id")
)
```

<div class="note">
**Reading it** — the trajectory through domain space. Disturbance typically
pushes points toward the origin; recovery is the return path. A return that
does not retrace its outbound path is hysteresis, and it is visible here as an
open loop.
</div>

<div class="caution">
**What it does not show** — the axes are domain composite scores, not the
hidden states of Section 3. Physiology is not Kinetics; Soil is not Capacity.
Relabelling them as mechanistic properties would be a category error.
</div>

# Recovery signatures

```{r recovery}
rec <- rri_recovery_metrics(
  res           = res,
  id            = sim$id,
  time_col      = "time",
  group_cols    = c("plot", "depth", "plant_id"),
  perturb_start = PERTURB_START,
  perturb_end   = PERTURB_END,
  rri_col       = "RRI"
)

rec[1:4, c("plot", "depth", "plant_id", "baseline_rri", "depth_min_frac",
           "tau_lag", "overshoot_frac", "incomplete_return_frac",
           "displaced_plateau_flag", "fit_status")]
```

<div class="note">
**Reading it** — one row per trajectory. `depth_min_frac` is how far the score
fell relative to baseline; `tau_lag` is how long recovery took to begin;
`incomplete_return_frac` is the terminal shortfall. `fit_status` reports
whether the rate estimate is trustworthy — always read `k` together with it.
</div>

<div class="caution">
**What it does not show** — `alt_routing_flag` is retained as `NA` on purpose.
A displaced plateau is consistent with alternative electron routing but does
not establish it, so the package refuses to claim otherwise. Use
`displaced_plateau_flag` and describe it as a displacement.
</div>

## Recovery map across all trajectories

```{r recovery-map, fig.height=4.8}
plot_rri_recovery_map(
  res           = res,
  id            = sim$id,
  rec           = rec,
  time_col      = "time",
  group_cols    = c("plot", "depth", "plant_id"),
  perturb_start = PERTURB_START,
  perturb_end   = PERTURB_END
)
```

<div class="note">
**Reading it** — one row per trajectory, colour = RRI through time. Scan
vertically at any time point to compare units; scan horizontally to follow one
unit. Rows that stay dark to the right of the disturbance band did not
recover.
</div>

## Ranking trajectories by signature

```{r landscape, eval=requireNamespace("tidyr", quietly=TRUE) && requireNamespace("tidyselect", quietly=TRUE), fig.height=5}
## Name the metrics explicitly rather than relying on the function default.
## Older HRRI builds defaulted to A_norm / O_norm / tau_r, which
## rri_recovery_metrics() no longer produces; being explicit makes this chunk
## work against either version and documents which signatures are shown.
plot_rri_recovery_landscape(
  rec,
  metrics = intersect(
    c("depth_min_frac", "overshoot_frac", "I_norm", "k", "tau_lag", "t_half"),
    names(rec)
  ),
  order_by = "I_norm"
)
```

<div class="note">
**Reading it** — trajectories as rows, recovery signatures as columns, each
column scaled within the cohort. It answers "which units behaved similarly, and
on which signature do they differ?" — the ordering is by incomplete return.
</div>

<div class="caution">
**What it does not show** — scaling is cohort-relative, so a "high" cell means
high *within this run*, not high in absolute terms. Two datasets cannot be
compared cell-by-cell.
</div>

# Property diagnostics

```{r properties, fig.height=5}
props <- rri_property_scores(res, rec = rec)
props$property_table

plot_rri_properties(props, rri_value = mean(scored$RRI, na.rm = TRUE))
```

<div class="note">
**Reading it** — the four diagnostics on one radar. Unavailable properties stay
missing rather than being imputed, so a short spoke means "not supported by the
supplied data", not "low".
</div>

<div class="caution">
**What it does not show** — these are *named after* the four controls but are
not measurements of them. Capacity here is an oxidative-oriented feature
composite; Connectivity is an association descriptor; Kinetics is a
recovery-speed descriptor; Memory is a persistent-displacement descriptor.
Check `props$property_table` to see the method behind each score.
</div>

# Did HRRI recover the hidden state?

Because the simulator defines a target, we can check agreement directly. This
is an internal consistency check, not empirical validation.

```{r validation, fig.height=4}
ok <- is.finite(scored$RRI) & is.finite(sim$latent_truth)
r  <- stats::cor(scored$RRI[ok], sim$latent_truth[ok])
cat("r(RRI, latent_truth) =", round(r, 3), "on", sum(ok), "observations\n")

ggplot(data.frame(truth = sim$latent_truth[ok], RRI = scored$RRI[ok]),
       aes(truth, RRI)) +
  geom_point(alpha = 0.25, size = 1.1, colour = "#2f6b6b") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "#8c4a2f", fill = "#8c4a2f", alpha = 0.12) +
  labs(x = "Simulator target  z(t)", y = "HRRI score",
       title = sprintf("Agreement with the prescribed target (r = %.3f)", r))
```

<div class="caution">
**What it does not show** — this is agreement with a target *we defined*. It
demonstrates that the scoring path is internally coherent. It is not predictive
accuracy, not held-out error, and not evidence that HRRI tracks resilience in
any real soil.
</div>

# Using your own data

Every function above accepts plain data frames. Replace the simulator with your
own measurements, keeping rows aligned across blocks:

```{r own-data, eval=FALSE}
res <- rri_pipeline(
  soil  = my_soil,      # Eh, pH, Fe pools, EAC/EDC ...
  plant = my_plant,     # SPAD, Fv/Fm, ROL ...
  micro = my_micro,     # ASV table or functional genes
  id    = my_ids,       # plot, depth, plant_id, time
  direction_anchor_soil = "Eh",
  direction_anchor_phys = "FvFm"
)
```

Missing a domain is fine — supply what you have. Absent domains stay `NA`,
remaining weights renormalise per row, and coverage is reported so a reduced
panel is never silently treated as a complete one.

<div class="method">
A reduced panel changes the estimand. Scores from a soil-only run and a
three-domain run are not interchangeable; compare them through
`rri_sensitivity()` rather than assuming equivalence.
</div>

# Session information

```{r session}
sessionInfo()
```
